---
title: "EDA: Numeric Summaries"
output: github_document
---


```{r}
library(tidyverse)
library(ggridges)
```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
       USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
  month = lubridate::floor_date(date, unit = "month")) |> 
  select(name, id, everything())
```

## Initial numeric work

```{r}
weather_df |> 
  ggplot(aes(x = prcp)) +
  geom_histogram()
```

here are the big outliers

```{r}
weather_df |> 
  filter(prcp > 1000)
```

```{r}
weather_df |> 
  filter(tmax >= 20, tmax <= 30) |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point()
```

## Grouping 

```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(n_obs = n())
```

```{r}
weather_df |> 
  group_by(month) |> 
   summarize(n_obs = n())
```

```{r}
weather_df |> 
  count(name)
```

```{r}
weather_df |> 
  count(name, month) |> 
  pivot_wider(
    names_from = name,
    values_from = n
  )
```

## General summaries

```{r}
weather_df |> 
  group_by(name) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
  )
```

taking mean on tmax
```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
  )
```


taking median and standard deviation 
```{r}
weather_df |> 
  group_by(name) |> 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmax = median(tmax, na.rm = TRUE),
    sd_tmax = sd(tmax, na.rm = TRUE)
  )
```


```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) |> 
  ggplot(aes(x = month, y= mean_tmax, color= name)) +
  geom_point() +
  geom_line()
```

how to pivot wider function and set it like an actual table rather than a data frame visual 
```{r}
weather_df |> 
  group_by(name, month) |> 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) |> 
  pivot_wider(
    names_from = name, 
    values_from = mean_tmax
  ) |> 
  knitr::kable(digits = 2)
```

## Grouped mutate 

```{r}
weather_df |>
  group_by(name) |>
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax) |> 
  ggplot(aes(x = date, y = centered_tmax, color = name)) + 
    geom_point() 
```

```{r}
weather_df |>
  group_by(name, month) |>
  mutate(temp_ranking = min_rank(tmax))
```

can also filter
```{r}
weather_df |>
  group_by(name, month) |>
  mutate(tmax_rank = min_rank(desc(tmax))) |> 
  filter(tmax_rank < 2 )
```

lags
used to compare an observation to it’s previous value. This is useful, for example, to find the day-by-day change in max temperature within each station over the year

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(yesterday_tmax = lag(tmax, 3))
```

This kind of variable might be used to quantify the day-by-day variability in max temperature, or to identify the largest one-day increase

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(temp_change = tmax - lag(tmax)) |> 
  summarize(
    sd_temp_change = sd(temp_change, na.rm = TRUE)
  )
```

