---
title: "viz_and_eda"
output: github_document
---

```{r}
library(tidyverse)
library(ggridges)
```

We’ll be working with NOAA weather data, which is downloaded using rnoaa::meteo_pull_monitors function in the code chunk below

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

let's make a plot!

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
    geom_point()
```

Pipes and stuff

```{r}
weather_df |> 
  filter(name == "CentralPark_NY") |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()

ggp_nyc_weather = 
  weather_df |> 
  filter(name == "CentralPark_NY") |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()

ggp_nyc_weather
```

## Fancy Plot 

```{r}
ggplot(weather_df, aes(x = tmin, y= tmax, color = name)) +
  geom_point(aes(color = name), alpha = 0.3) +
  geom_smooth(se= FALSE)

```

Plot with facets

```{r}
ggplot(weather_df, aes(x= tmin, y = tmax, color = name)) +
  geom_point(alpha=.3) +
    geom_smooth() +
    facet_grid(. ~ name)
```

let's try a different plotting, temp is boring (we will use precipation)

```{r}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

trying assigning a specific color
!= --> not equal

```{r}
weather_df |> 
  filter(name == "CentralPark_NY") |> 
  ggplot(aes(x = date, y = tmax)) +
  geom_point(alpha = .7, size = .5)
```




```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()
```

```{r}
weather_df |> 
  filter(name == "Molokai_HI") |> 
  ggplot(aes(x= date, y = tmax)) +
  geom_line(alpha = .5) +
  geom_point(size = .5)
```


## univariate plotting

histogram

```{r}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_histogram(position = "dodge", binwidth = 2)
```

let's use a density plot

```{r}
ggplot(weather_df, aes(x = tmax, fill = name)) + 
  geom_density(alpha = .4, adjust = .5, color = "blue")
```

using boxplots!!

```{r}
ggplot(weather_df, aes(x = name, y = tmax)) + geom_boxplot()
```

violin plots?

```{r}
ggplot(weather_df, aes(y = tmax, x = name)) + 
  geom_violin()
```

ridge plot

```{r}
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)
```

## saving and embedding plots

```{r}
ggp_weather = 
  weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point() 

ggsave("ggp_weather.pdf", ggp_weather, width = 8, height = 5)
```

```{r, fig.width = 6, fig.asp= .6}
ggp_weather
```


